---
title: "Heart Disease Project"
author: "Anna M"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

The Heart Disease Dataset from Kaggle will be used for exploratory data 
analysis (EDA). As this data set is already cleaned, EDA will focus on investigating 
and visualizing the attributes of this data set. This project will investigate 
whether there is a correlation between heart disease and other parameters in 
this data set.

```{r, include = FALSE}
##Packages downloaded (left in comments as CRAN specs are different for knitr)
##install.packages("corrplot")
##install.packages("ggplot2")
##install.packages("dplyr")
```


```{r load-packages, include = FALSE}
##Packages loaded
library(corrplot)
library(ggplot2)
library(dplyr)
```

Dataset downloaded

```{r}
heart <- read.csv("heart.csv")
```

Brief overview of correlation of variables 

```{r}
corrplot(cor(heart), type="upper")
```

Identify variables in dataset

```{r}
ls()
ls(heart)
```

Categorical variables changed from numeric values to string values

```{r}
heart$sex[heart$sex == 0] = "female"
heart$sex[heart$sex == 1] = "male"
heart$cp[heart$cp == 0] = "typical angina"
heart$cp[heart$cp == 1] = "atypical angina"
heart$cp[heart$cp == 2] = "non-anginal pain"
heart$cp[heart$cp == 3] = "asymptomatic"
##trestbps does not need to be changed as it shows resting blood pressure (unique)
##chol does not need to be changed as it shows cholesterol level (unique)
heart$fbs[heart$fbs == 0] = "blood sugar = or < 120mg/dl"
heart$fbs[heart$fbs == 1] = "blood sugar > 120mg/dl"
heart$restecg[heart$restecg == 0] = "normal ECG"
heart$restecg[heart$restecg == 1] = "ST-T wave abnormality"
heart$restecg[heart$restecg == 2] = "left ventricular hypertrophy"
##thalach does not need to be changed as it shows maximum heart rate (unique)
heart$exang[heart$exang == 0] = "no"
heart$exang[heart$exang == 1] = "yes"
##oldpeak does not need to be changed as it shows ST depression (unique)
heart$slope[heart$slope == 0] = "upsloping"
heart$slope[heart$slope == 1] = "flat"
heart$slope[heart$slope == 2] = "downsloping"
##ca does not need to be changed as it shows number of major vessels (numeric)
heart$thal[heart$thal == 1] = "normal"
heart$thal[heart$thal == 2] = "fixed defect"
heart$thal[heart$thal == 3] = "reversable defect"
```

We will use "target1" to keep have both the numeric value of target and the
categorical value of target1.

```{r}
heart$target1[heart$target == 0] = "no"
heart$target1[heart$target == 1] = "yes"
```

Q: What is proportion of sample population with heart disease?

```{r}
##Proportion table
prop.table(table(heart$target1))

##Proportion table with result rounded to two decimal places
round(prop.table(table(heart$target1)), 2)
```

Q: What subpopulation is heart disease most prevalent in? 

```{r}
##Distribution of sample
ggplot(heart, aes(x=age)) + 
  geom_histogram() + 
  ggtitle("Distribution of age of the population") + 
  xlab("Age") + 
  ylab("Density")
```

The distribution is almost normal, but also slightly skewed to the right. 
Looking at age groups may be helpful, as the primary age group of the sample is 50-65.

```{r}
##Min and max age ranges
min(heart$age)
max(heart$age)
```

We will look at the range of ages from 29-77. Now we separate age groups into 
4-year sections.

```{r}
##The cut function was not including lowest group (29-33) and instead putting
##putting this value at the end of the list with NA, but this was fixed by adding
#the last argument.
heart$age_group = cut(heart$age, breaks = seq(29,77,4), include.lowest = TRUE)
```

Now, we can find the number of people with heart disease for each age group.

```{r}
##NB: we are using target here (numeric), not target1 (character) because 
##summarise cannot take character vectors. 

HD_by_age = heart %>%
  group_by(age_group) %>%
  summarise(heart_disease = sum(target))

HD_by_age
```

We can visualise this with a barplot.

```{r}
HD_by_age %>%
  ggplot(aes(x = age_group, y = heart_disease)) + 
  geom_bar(stat = "identity", fill = "red", alpha = .5, width = .4) + 
  xlab("Age Groups") + 
  ylab("# People with HD") + 
  ggtitle("# HD by Age Group") + 
  theme_bw()
```

This is a good visualisation, but does not show the relative rate of HD.
We need to find the proportion of people with HD by age group, as our sample
does not include as many people with ages below 30 or above 70. 

```{r}
prop_in_age <- heart %>%
  group_by(age_group) %>%
  summarise(HD_prop = round(sum(target)/n(), 3)*100)

prop_in_age
```

We can visualise this with a barplot.

```{r}
prop_in_age %>%
  ggplot(aes(x = age_group, y = HD_prop)) + 
  geom_bar(stat = "identity", fill = "blue", alpha = .5, width = .4) + 
  xlab("Age Groups") + 
  ylab("Prop People with HD") + 
  ggtitle("Prop HD by Age Group") + 
  theme_bw()
```

100% of people 29-33 have HD. This is not representative of real life, so this
dataset is not a representative sample. Thus, we cannot infer anything about 
how age contributes to HD from this dataset.

#### Exploring Variable Relationships: Sex</font> 

Q: What is proportion of sample population with HD by sex?

```{r}
prop.table(table(heart$sex))
round(prop.table(table(heart$sex)), 2)
```

Q: What is the proportion of males vs females with HD?

```{r}
round(prop.table(table(heart$sex, heart$target1)), 2)
```

We find that the female proportion of HD is much higher than male. This 
could be due to females having a lower sample overall, indicating a slight
bias towards a sample of more women who have HD. 

#### Exploring Variable Relationships: Slope of Peak Exercise</font> 

First, I need to understand what this variable represents. 

An ST slope shifts relative to exercise-induced increments in HR. 
Importance: The ST/HR slope has been proposed as a more accurate ECG 
criterion for diagnosing significant coronary artery disease. 

* Upwards slope = normal ST during exercise
* Flat/downwards slope = may indicate coronary ischemia

To see this represented in our sample data, we will look at how ST slope
relates to HR incidence. Since we are using yes/no, target1 (categorical) 
was used. To get the barplot side by side instead of stacked, dodge was used.

* slope = 0 is upwards
* slope = 1 is flat
* slope = 2 is downwards 

```{r}
ggplot(heart, aes(x = slope, fill = target1)) + 
  geom_bar(position = "dodge") + 
  xlab("Slope") + 
  ylab("Count") + 
  ggtitle("HD Incidence by ST Slope") + 
  scale_fill_discrete(name = "Heart disease", labels = c("No", "Yes"))
```

From this, we see that those with downsloping ST slopes are most likely to 
have higher changes of HD compared with flat or upwards slope. 

Now, let's look at the above proportions by sex.

```{r}
male_data = heart[heart$sex=="male",]
female_data = heart[heart$sex=="female",]

##The comma at the end (row, column) means that we are looking at rows 
##have "male" in all the entire column (hence it is left blank)
```

Let's visualise this with a barplot (MALES).

```{r}
ggplot(male_data, aes(x = slope, fill = target1)) + 
  geom_bar(position = "dodge") + 
  xlab("Slope") + 
  ylab("Count") + 
  ggtitle("HD Incidence of ST Slope - Males") + 
  scale_fill_discrete(name = "HD", labels = c("No", "Yes"))
```

Let's visualise this with a barplot (FEMALES). 

```{r}
ggplot(female_data, aes(x = slope, fill = target1)) + 
  geom_bar(position = "dodge") + 
  xlab("Slope") + 
  ylab("Count") + 
  ggtitle("HD Incidence of ST Slope - Females") + 
  scale_fill_discrete(name = "HD", labels = c("No", "Yes"))
```

From this, we see that the proportion of women who have downsloping ST is much 
higher than the males proportion for this group. Additionally, compared to the 
original graph (not grouped by sex), the male trend follows this original graph 
whereas the female population is very different. As the female population is 
much less, it shifts the overall data pattern less, so it makes sense the overall
data reflects the male pattern of HD incidence by type more. 

#### Exploring Variable Relationships: Number of Major Vessels</font> 

We can use a mosaic plot to visualise the differences in HD by number of 
major vessels. 

```{r}
mosaicplot(table(heart$target1, heart$ca), col = c("red", "yellow", "blue", "black", "pink"), 
           main = "HD for Major Vessels")
```

We see that almost double the amount of people who have no major vessel
have heart disease. In general, very few people have 4 major vessels. 

We can visualise this by sex.

```{r}
mosaicplot(table(male_data$target1, male_data$ca), col = c("red", "yellow", "blue", "black", "pink"), 
           main = "HD for Major Vessels - Male")

mosaicplot(table(female_data$target1, female_data$ca), col = c("red", "yellow", "blue", "black", "pink"), 
           main = "HD for Major Vessels - Female")
```

We find that both male and female subpopulations reflect the overall data.

#### Exploring Variable Relationships: ST Depression by Exercise</font> 

First, we will look at the distribution of ST depression by HD incidence. 

```{r}
ggplot(heart, aes(x = target1, y = oldpeak)) + 
  geom_boxplot()
```

We see that ST depression for HD incidence is higher (eg IGQ) than that
for no HR incidence

Q: Does this depression change by age? Does this have an impact on HD? 

```{r}
ggplot(heart, aes(x = age, y = oldpeak, color = target1)) + 
  geom_point(alpha = 0.3) + 
  labs(color = "Heart Disease State") + 
  xlab("Age") + 
  ylab("ST Depression") + 
  ggtitle("Age vs Resting Blood Pressure Separated by Heart Condition") +
  guides(size = FALSE)
```

We see that there is no clear relationship ebtween age and ST depresison. 

#### Exploring Variable Relationships: Resting Blood Pressure</font> 

```{r}
ggplot(heart, aes(x = target1, y = trestbps)) + 
  geom_boxplot() + 
  xlab("Heart Disease State") + 
  ylab("Resting BP") + 
  ggtitle("Boxplots of Resting BP by Heart Condition")
```

Medians of these boxplots are similar. We can interrogate this relationship
further by looking at ST Depression.

```{r}
ggplot(heart, aes(x = age, y = trestbps, color = target1, size = factor(oldpeak))) + 
  geom_point(alpha = 0.3) + 
  xlab("Heart Disease State") + 
  ylab("Resting BP") + 
  ggtitle("Boxplots of Resting BP by Heart Condition") + 
  guides(size = FALSE)
```

From this, we see that resting blood pressure is low (< 100), heart disease
cases are higher than no heart disease cases. Additionally, most dots are red
meaning that more people with ST depression have no heart disease. And, 
more large dots are in the higher age range so ST Depression is higher in 
older populations. 